---
title: "Transcriptomics_Homework-2_Charlee-Cobb_cac9995"
author: "Charlee Cobb"
date: "2023-02-20"
output: pdf_document
---
**Part 1: Answer the following questions. (20 pts)**
**__________________________________________________________________**
***1. Why is it important to know the variance and standard error of a statistic***
***for sample data? How do we calculate these values in R?***
Understanding variance in data is important because it shows the difference a 
given point has from the mean. This would show if our data had many points that
are outliers or how spread out our data is. Standard error compares the mean of
a sample to the mean of the population, and the value helps us decide how 
representative a sample is for the population. To calculate the standard error
in R by getting the value of standard deviation with the function sd(), and 
dividing that value by the square root of the number of samples. To get variance
in R, we can use the var() function on the sample dataset.


***2. What are the differences between the varieties of two-sample t-tests,***
***and how can we tell which to use and how do we tell R how to apply***
***the correct test using t-test()?***
Paired t-test and unpaired t-tests are the varieties of two-sample t-tests. Paired
t-tests look at two samples that are related to each other. For example, in the 
paired t-test we might compare two different observations of one cancer tumor in
different from different time points, pretreatmnt and after treatment. With 
Unpaired t-tests, the two samples are completely independent from each other. 
In R, you use the t.test() function, and set the flag 'paired' to TRUE for a 
paired t-test, and FALSE for an unpaired t-test.


***3. What is the R function for a nonparametric equivalent to a two-sample t-test?***
The wilcoxon test using the wilcox_test() function.


***4. What statistic is more appropriate when you want to identify***
***interactions between multiple factors?***
ANOVA statistics are better to use when checking interactions between multiple
factors.


***5. When is the lm() function more appropriate to use than aov()***
***function to perform an anova analysis?***
aov() looks at variation within groups and variation between groups without respect
to each other. lm() looks for the slope made in the data, and looks at each slope separately.


**Part 2: We will be working with a publicly available dataset: GSE63741.**
**The experiment compares expression of a set of genes during different skin**
**diseases. Our goal is to identify the genes that are significantly different**
**from the control (healthy) samples. (80pts)**
**__________________________________________________________________**
***1. Write a function getPvalue that performs a linear model lm() to***
***compare the control samples to the different skin diseases. (40pts)***
***- The function should have two inputs***
    ***1: numerical vector of all values for one gene,***
    ***2: a factor that specifies which column belongs to which group.***
***- The function should have one output ( a vector of p-values for***
***each disease coefficient.***
```{r}
#getPvalue fucntion

getPvalue <- function(vect, fact){
  gene <- unlist(vect, use.names = FALSE)
  p <- lm(gene ~ fact)$coefficients
  return(p)
}
```


***2. Use the getPvalue function in a loop or apply function to retrieve the***
***pvalues for each gene. (10pts)***
```{r}
#load data sets
GSE63741 <- read.delim("GSE63741_series_matrix.txt")
sample <- read.delim("sampletype.txt")
#head(GSE63741) #< used to check object
#head(sample) #< used to check object

#make factor of groups
group_names <- unlist(sample[1,], use.names = FALSE)
#group_names #< used to check object
group_factor <- factor(group_names)
group_factor #< used to check object

#making a test run to help model the function
testing <- lm(unlist(GSE63741[1, 2:151], use.names = FALSE) ~ group_factor)$coefficients

#I tried to use the function I worte above, but kept running into bugs when I 
#ran this:
#res <- apply(GSE63741, 1, getPvalue(vect = GSE63741[2:151], fact=group_factor) )


#So I wrote the function in-line with the apply function
res <- apply(GSE63741, 1, function(x) lm(x[2:151]~group_factor)$coefficients)
head(res)
hist(res)
sum(res < -0.05)
```



***3. Adjust the p-values using the FDR method. (20pts)***
```{r}
adjusted_pvalue <- p.adjust(res, method = "fdr")
```



***4. Gene regulation: How many genes have an adjusted p-value < 0.05 in at***
***least one skin disease? (10pts)***
in total, 4336 genes had an adjusted p value above 0.05. In Eczema alone, there
were 869 genes that have an adjusted p-value over 0.05
```{r}
sum(adjusted_pvalue < 0.05)

eczema <- res[2,]
adjusted_eczema <-p.adjust(eczema, method = "fdr")
sum(adjusted_eczema < 0.05)
```




unrelated test:
```{r}

#subset data into control and treatment dataframes
control_df <- GSE63741[ ,2:32]
treated_df <- GSE63741[ ,33:151]

head(control_df)
head(treated_df)

#calculate the means of each row, then save those means in a vector
control_means <- rowMeans(control_df)
treated_means <- rowMeans(treated_df)

#testing to see lm between vector of means
head(control_means)
lm(control_means ~ treated_means)
```
